%*******************************************************
% Acknowledgements
%*******************************************************


\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter{Acknowledgements}
\label{chap:acknowledgments}
\mtcaddchapter
I owe profound thanks to several people for the following
work. 

Firstly, I thank my supervisor, dr. Toon Van Hal, who from the get-go
demonstrated a very open-minded attitude in the face of unorthodox
subject matter for a classical philology thesis. He has demonstrated
exemplary patience with the stop-start rhythm of development of this
work. His corrections, suggestions and occasional nudges (usually
delivered electronically and with utmost tact) were an invaluable help
and encouragement. It is evident that without him, this work would not
have been possible.

Next, I am grateful to dr. Francis Maes, until recently a postgraduate
student at the Department of Computer Science at the KU Leuven Faculty
of Science, who during a conversation on this thesis directed me to a
set of state-of-the-art papers on machine learning for natural
language processing. His suggestions form the methodological bedrock
of this work in its current form.

I would also like to extend my thanks to my godmother Mar\'ia
Aguililla, who immediately accepted to proofread this work and did so
diligently, as well as Erwin De Koster for temporarily providing me
with computational power in the form of an iMac.

I offer my gratitude to my high school classics teacher, Guy van den
Heule, whose passion for Greco-Roman language and culture combined
with a lively and engaging teaching style pushed me to pursue the
study of the classical languages at a higher level.

I would also like to thank all my friends who offered me support and lended
me their proverbial ear in some way when the going got rough or showed
more than a cursory interest in this work, most of whom definitely
know who they are; they are too many to mention and I fear forgetting
but one.

I have not mentioned the most important persons of all: my parents,
who have been unflagging in their encouragement, interest and
support. They have enabled me to continue work on this thesis and to
pursue my study of computer science, which has granted me the
technical knowledge necessary to complete this work. To them I owe the
greatest thanks of all.

Finally, I invoke the memory of my good friend Lorena Lage Pi\~neiro,
who would have been mentioned in the previous paragraphs, if not for
the intervention of fate. Her sudden passing a week before this
writing deeply saddened me, and it is to her that I dedicate this
work.

\endgroup



